visitation_bias

### 2021-04-20 code/ms2_ktype_preference_nullmodel_analysis.Rmd
switched to proportional visitation data for null models

---
title: "visitation_bias"
author: "Doug Sponsler"
date: "12/14/2020"
output: html_document
---
# Preparations
## Packages 
```{r, warning=FALSE, include=FALSE, echo = FALSE}
library(bipartite)
library(econullnetr)
library(patchwork)
library(mgcv)
library(mgcViz)
library(tidymv)
library(tidyverse)
library(lubridate)
library(vegan)
library(ggvegan)
library(ggrepel)
```

@@

visit_cover <- read_csv("../data/processed_data/visit_cover.csv")
```

# Null model 1: bb.sp x ktype, fully pooled
```{r}
# The "consumers" table input to generate_null_net()
con <- prop_visits_ktype %>%
  unite(id, c(site, date)) %>%
  select(id, bb.sp, ktype, visits) %>%
  pivot_wider(values_from = visits, names_from = ktype, values_fill = 0) %>%
  select(-c("0", "10", "9"))

@@

# Generate null model
null1 <- generate_null_net(consumers = con[, 2:9],
                           resources = res[, 2:8],
                           sims = 5000, # raised from default of 1000 (takes a while to run)
                           data.type = "counts",
                           c.samples = con[, 1], # parsed by sample
                           r.samples = res[, 1])

#saveRDS(null1, file = "../output/nullmodel_ktype.rds")
null1 <- readRDS("../output/nullmodel_ktype.rds")

# Tabulate output and generate tests
null_tab1 <- test_interactions(null1, signif.level = 0.99) %>% # make it a bit more conservative to account for multiple comparisons
  as_tibble() %>%
  select(bb.sp = Consumer, ktype = Resource, everything()) %>%
  group_by(bb.sp) %>%
  mutate(total.abund = sum(Observed), # total abundance per species
         preference = Observed - Null, # a working definition of preference
         preference.upper99 = Observed - Upper.99.CL, 
         preference.lower99 = Observed - Lower.99.CL,

         # If preference equals the difference between # observed visits and the null model prediction, then dividing preference by the total abundance of each species yields a standardized preference. Right? 
         preference.std = preference/total.abund, 
         Observed.std = Observed/total.abund,
         Upper.99.CL.std = Upper.99.CL/total.abund,
         Lower.99.CL.std = Lower.99.CL/total.abund) %>%
  left_join(bb_traits) %>%
  ungroup() %>%
  mutate(bb.sp = fct_reorder(factor(bb.sp), pbl.w))
```

@@

ggsave("../output/null_model_ktypeXtongueclass_preference.pdf", height = 6, width = 7)
ggsave("../output/null_model_ktypeXtongueclass_preference.png", height = 6, width = 7)
```

@@

# Null model 2: bb.sp x ktype, by site
```{r}
# Nest and join consumer and resource data frames
con_nest <- prop_visits_ktype %>%
  select(site, date, bb.sp, ktype, visits) %>%
  pivot_wider(values_from = visits, names_from = ktype, values_fill = 0) %>%
  select(-c("0", "10", "9")) %>%
  group_by(site) %>%
  nest() %>%
  rename(bees = data)

@@

## RUN ON HPC!
# sitewise_null_models <- conres %>%
#   
#   # Generate null models
#   mutate(null.model = map2(bees, plants, function(x, y) 
#     generate_null_net(consumers = x[, 2:9],
#                       resources = y[, 2:8],
#                       sims = 2000,
#                       data.type = "counts",
#                       c.samples = x[, 1],
#                       r.samples = y[, 1]))) %>%
#   
#   # Tabulate results (with test)
#   mutate(null.model.tab = map(null.model, function(x) 
#     test_interactions(x, signif.level = 0.99) %>%
#       as_tibble() %>%
#       mutate(preference = Observed - Null,
#              preference.upper99 = Observed - Upper.99.CL,
#              preference.lower99 = Observed - Lower.99.CL) %>%
#       select(bb.sp = Consumer, ktype = Resource, everything()))) %>%
#   
#   # Unnest and add site data and bumble bee trait data
#   unnest(null.model.tab) %>%
#   left_join(site_data) %>%
#   left_join(bb_traits) %>%
#   
#   # Calculate preference/total abundance per species
#   group_by(site, bb.sp) %>%
#   mutate(total.abund = sum(Observed)) %>%
#   ungroup() %>%
#   mutate(preference.std = preference/total.abund)

#write_rds(sitewise_null_models, "../output/nullmodel_ktype_sitewise.rds")
#write_rds(sitewise_null_models, "./nullmodel_ktype_sitewise.rds") # hpc
sitewise_null_models <- readRDS("../output/nullmodel_ktype_sitewise.rds")

sitewise_null_models <- sitewise_null_models %>%
  filter(total.abund > 0) %>% # for each species, drop sites where the species did not occur
  unite(bb.ktype, c(bb.sp, ktype), remove = FALSE) %>%
  mutate(ktype = factor(ktype),
         bb.sp = factor(bb.sp),
         bb.ktype = factor(bb.ktype))
```

## Diet similarity vs. preference similarity
```{r}
prefs2 <- sitewise_null_models %>%
  select(site, bb.sp, ktype, preference.std) %>%
  unite(site.bb.sp, c(site, bb.sp)) %>%
  pivot_wider(names_from = ktype, values_from = preference.std) %>%
  column_to_rownames("site.bb.sp")

## Visualize site model
```{r}
# ktype preference x bb.sp ~ elevation
ggplot(sitewise_null_models, aes(elev.mean, preference.std, color = ktype)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
  facet_wrap(~bb.sp)

@@

## Overall preference of each ktype ~ elevation
```{r}

# Null model 3: bb.sp x plant genus, fully pooled
```{r}
# The "consumers" table input to generate_null_net()
con3 <- visit_cover %>%
  group_by(site, date, bb.sp, plant.genus) %>%
  summarize(visits = sum(visits)) %>%
  unite(id, c(site, date)) %>%
  select(id, bb.sp, plant.genus, visits) %>%
  pivot_wider(values_from = visits, names_from = plant.genus, values_fill = 0)

# The "resources" table input to generate_null_net()
res3 <- visit_cover %>%
  group_by(site, date, bb.sp, plant.genus) %>%
  summarize(flower.cover = sum(flower.cover)) %>%
  group_by(site, date) %>%
  mutate(perc.flower.cover = flower.cover/sum(flower.cover)) %>%
  unite(id, c(site, date)) %>%
  select(id, plant.genus, perc.flower.cover) %>%
  distinct() %>%
  pivot_wider(values_from = perc.flower.cover, names_from = plant.genus, values_fill = 0)

# Generate null model
null3 <- generate_null_net(consumers = con3[, 2:201],
                           resources = res3[, 2:200],
                           sims = 5000, # raised from default of 1000 (takes a while to run)
                           data.type = "counts",
                           c.samples = con3[, 1], # parsed by sample
                           r.samples = res3[, 1])

saveRDS(null3, file = "./nullmodel_plant.genus.rds") # hpc
null3 <- readRDS("../output/nullmodel_ktype.rds")

# Tabulate output and generate tests
null_tab1 <- test_interactions(null1, signif.level = 0.99) %>% # make it a bit more conservative to account for multiple comparisons
  as_tibble() %>%
  select(bb.sp = Consumer, ktype = Resource, everything()) %>%
  group_by(bb.sp) %>%
  mutate(total.abund = sum(Observed), # total abundance per species
         preference = Observed - Null, # a working definition of preference
         preference.upper99 = Observed - Upper.99.CL, 
         preference.lower99 = Observed - Lower.99.CL,

         # If preference equals the difference between # observed visits and the null model prediction, then dividing preference by the total abundance of each species yields a standardized preference. Right? 
         preference.std = preference/total.abund, 
         Observed.std = Observed/total.abund,
         Upper.99.CL.std = Upper.99.CL/total.abund,
         Lower.99.CL.std = Lower.99.CL/total.abund) %>%
  left_join(bb_traits) %>%
  ungroup() %>%
  mutate(bb.sp = fct_reorder(factor(bb.sp), pbl.w))
```

# Null model 4: bb.sp x plant genus, fully pooled
```{r}
# The "consumers" table input to generate_null_net()
con4 <- visit_cover %>%
  group_by(site, date, bb.sp, plant.family) %>%
  summarize(visits = sum(visits)) %>%
  unite(id, c(site, date)) %>%
  select(id, bb.sp, plant.family, visits) %>%
  pivot_wider(values_from = visits, names_from = plant.family, values_fill = 0)

# The "resources" table input to generate_null_net()
res4 <- visit_cover %>%
  group_by(site, date, bb.sp, plant.family) %>%
  summarize(flower.cover = sum(flower.cover)) %>%
  group_by(site, date) %>%
  mutate(perc.flower.cover = flower.cover/sum(flower.cover)) %>%
  unite(id, c(site, date)) %>%
  select(id, plant.family, perc.flower.cover) %>%
  distinct() %>%
  pivot_wider(values_from = perc.flower.cover, names_from = plant.family, values_fill = 0)

# Generate null model
null4 <- generate_null_net(consumers = con4[, 2:56],
                           resources = res4[, 2:55],
                           sims = 5000, # raised from default of 1000 (takes a while to run)
                           data.type = "counts",
                           c.samples = con4[, 1], # parsed by sample
                           r.samples = res4[, 1])

saveRDS(null4, file = "./nullmodel_plant.family.rds") # hpc
null3 <- readRDS("../output/nullmodel_ktype.rds")

# Tabulate output and generate tests
null_tab1 <- test_interactions(null1, signif.level = 0.99) %>% # make it a bit more conservative to account for multiple comparisons
  as_tibble() %>%
  select(bb.sp = Consumer, ktype = Resource, everything()) %>%
  group_by(bb.sp) %>%
  mutate(total.abund = sum(Observed), # total abundance per species
         preference = Observed - Null, # a working definition of preference
         preference.upper99 = Observed - Upper.99.CL, 
         preference.lower99 = Observed - Lower.99.CL,

         # If preference equals the difference between # observed visits and the null model prediction, then dividing preference by the total abundance of each species yields a standardized preference. Right? 
         preference.std = preference/total.abund, 
         Observed.std = Observed/total.abund,
         Upper.99.CL.std = Upper.99.CL/total.abund,
         Lower.99.CL.std = Lower.99.CL/total.abund) %>%
  left_join(bb_traits) %>%
  ungroup() %>%
  mutate(bb.sp = fct_reorder(factor(bb.sp), pbl.w))
```

>>>>

---
title: "visitation_bias"
author: "Doug Sponsler"
date: "12/14/2020"
output: html_document
---
# Preparations
## Packages 
```{r, warning=FALSE, include=FALSE, echo = FALSE}
library(econullnetr)
library(patchwork)
library(mgcv)
library(mgcViz)
library(tidymv)
library(tidyverse)
library(lubridate)
library(vegan)
library(ggvegan)
library(ggrepel)
```

@@

# For setting up the HPC
# prop_visits_ktype <- read_csv("./prop_visits_ktype.csv") %>%
#   mutate(ktype = factor(ktype), # recode variables
#          tongue = factor(tongue),
#          visits = as.integer(visits))
# 
# prop_visits_ktype_pooled <- read_csv("./prop_visits_ktype_pool.csv") %>%
#   mutate(ktype = factor(ktype), # recode variables
#          visits = as.integer(visits))
# 
# bb_traits <- read_csv("./bb_traits.csv") %>%
#   select(bb.sp, pbl.w, tongue = pbl.w.class2)
# 
# site_data <- read_csv("./site_data.csv") %>%
#   dplyr::select(-c(slope.calc, slope.est, elev.min, elev.max, temp.mean, elev.class, elev.class2)) %>%
#   mutate(site = factor(site),
#          tree_line = factor(tree_line))
```

# Null model 1: bb.sp x ktype, fully pooled
```{r}
# The "consumers" table input to generate_null_net()
con <- prop_visits_ktype %>%
  unite(id, c(site, date)) %>%
  select(id, bb.sp, ktype, prop.visits) %>%
  pivot_wider(values_from = prop.visits, names_from = ktype, values_fill = 0) %>%
  select(-c("0", "10", "9"))

@@

# Generate null model
null1 <- generate_null_net(consumers = con[, 2:9],
                           resources = res[, 2:8],
                           sims = 1000, # raised from default of 100 (takes a while to run)
                           data.type = "quantities", # since we are using proportional visitation
                           c.samples = con[, 1], # parsed by sample
                           r.samples = res[, 1])

saveRDS(null1, file = "./null1.rds")
null1 <- readRDS("../output/null1.rds")

# Tabulate output and generate tests
null_tab1 <- test_interactions(null1, signif.level = 0.99) %>% # make it a bit more conservative to account for multiple comparisons
  as_tibble() %>%
  select(bb.sp = Consumer, ktype = Resource, everything()) %>%
  group_by(bb.sp) %>%
  mutate(preference = Observed - Null, # a working definition of preference
         preference.upper99 = Observed - Upper.99.CL, 
         preference.lower99 = Observed - Lower.99.CL) %>%
  left_join(bb_traits) %>%
  ungroup() %>%
  mutate(bb.sp = fct_reorder(factor(bb.sp), pbl.w))
```

@@

ggsave("../output/null_model_ktypeXtongueclass_preference.pdf", height = 6, width = 7)
ggsave("../output/null_model_ktypeXtongueclass_preference.png", height = 6, width = 7)

ggplot(null_tab1, aes(pbl.w, SES)) +
  geom_point(aes(color = Test)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
  scale_color_manual(values = c("gray40", "#F8766D", "#00B6EB")) +
  facet_wrap(~ktype) +
  labs(x = "Tongue length (mm)", y = "SES", 
       color = "Test (p < 0.01)") +
  theme_light()

ggplot(null_tab1B, aes(pbl.w, preference)) +
  geom_point(aes(color = Test)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
  scale_color_manual(values = c("gray40", "#F8766D", "#00B6EB")) +
  facet_wrap(~ktype) +
  labs(x = "Tongue length (mm)", y = "Standardized preference", 
       color = "Test (p < 0.01)") +
  theme_light()
```

@@

# Null model 2: bb.sp x ktype, by site
```{r}
# Nest and join consumer and resource data frames
con_nest <- prop_visits_ktype %>%
  select(site, date, bb.sp, ktype, prop.visits) %>%
  pivot_wider(values_from = prop.visits, names_from = ktype, values_fill = 0) %>%
  select(-c("0", "10", "9")) %>%
  group_by(site) %>%
  nest() %>%
  rename(bees = data)

@@

## RUN ON HPC!
null2 <- conres %>%

  # Generate null models
  mutate(null.model = map2(bees, plants, function(x, y)
    generate_null_net(consumers = x[, 2:9],
                      resources = y[, 2:8],
                      sims = 1000, # raised from default of 100
                      data.type = "quantities",
                      c.samples = x[, 1],
                      r.samples = y[, 1])))

#write_rds(null2, "../output/nullmodel_ktype_sitewise.rds")
write_rds(null2, "./null2.rds") # hpc

# 
#   # Tabulate results (with test)
#   mutate(null.model.tab = map(null.model, function(x)
#     test_interactions(x, signif.level = 0.99) %>%
#       as_tibble() %>%
#       mutate(preference = Observed - Null,
#              preference.upper99 = Observed - Upper.99.CL,
#              preference.lower99 = Observed - Lower.99.CL) %>%
#       select(bb.sp = Consumer, ktype = Resource, everything()))) %>%
# 
#   # Unnest and add site data and bumble bee trait data
#   unnest(null.model.tab) %>%
#   left_join(site_data) %>%
#   left_join(bb_traits) %>%
#   filter(total.abund > 0) %>% # for each species, drop sites where the species did not occur
#   unite(bb.ktype, c(bb.sp, ktype), remove = FALSE)
# 
# 
# 
# sitewise_null_models <- readRDS("../output/nullmodel_ktype_sitewise.rds") %>%
#   mutate(ktype = factor(ktype),
#          bb.sp = factor(bb.sp),
#          bb.ktype = factor(bb.ktype))
```

## Visualize site model
```{r}
ggplot(sitewise_null_models, aes(pbl.w, preference.std)) +
  geom_point(aes(color = Test)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
  scale_color_manual(values = c("gray40", "#F8766D", "#00B6EB")) +
  facet_wrap(~ktype) +
  labs(x = "Tongue length (mm)", y = "Standardized preference", 
       color = "Test (p < 0.01)") +
  theme_light()

ggplot(sitewise_null_models, aes(pbl.w, SES)) +
  geom_point(aes(color = Test)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
  scale_color_manual(values = c("gray40", "#F8766D", "#00B6EB")) +
  facet_wrap(~ktype) +
  labs(x = "Tongue length (mm)", y = "Standardized preference", 
       color = "Test (p < 0.01)") +
  theme_light()

# ktype preference x bb.sp ~ elevation
ggplot(sitewise_null_models, aes(elev.mean, preference.std, color = ktype)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
  facet_wrap(~bb.sp)

@@

# Diet similarity vs. preference similarity
```{r}
diet_similarity <- read_csv("../output/niche_overlap.csv") %>%
  select(site, bb.sp1, bb.sp2, diet.similarity = niche.overlap, level) 
# %>%
#   pivot_wider(values_from = niche.overlap, names_from = level) %>%
#   rename(genus.overlap = genus, family.overlap = family, morphotype.overlap = morphotype)

pref_proc <- function(x) {
  x %>%
    pivot_wider(names_from = ktype, values_from = preference.std) %>%
    column_to_rownames("bb.sp") %>%
    vegdist(method = "euclidean") %>%
    broom::tidy() %>%
    select(bb.sp1 = 1, bb.sp2 = 2, distance)
}

preference_similarity_sitewise <- sitewise_null_models %>%
  select(site, bb.sp, ktype, preference.std) %>%
  group_by(site) %>%
  nest() %>%
  mutate(pdif = map(data, pref_proc)) %>%
  unnest(pdif) %>%
  left_join(bb_traits, by = c("bb.sp1" = "bb.sp")) %>%
  select(bb.sp1, bb.sp2, distance, pbl.w1 = pbl.w) %>%
  left_join(bb_traits, by = c("bb.sp2" = "bb.sp")) %>%
  select(bb.sp1, bb.sp2, distance, pbl.w1, pbl.w2 = pbl.w) %>%
  mutate(tongue.diff = abs(pbl.w1 - pbl.w2)) %>%
  ungroup() %>%
  mutate(distance.std = distance/max(distance),
         pref.similarity = 1-distance.std)

diet_preference <- full_join(preference_similarity_sitewise, diet_similarity) %>%
  filter(!is.na(pref.similarity)) # the data set that was used to generate the niche overlap data includes slightly more data than the one used to generate the preference data due to a filtering step during the alignment of visitation data to floral. Only 10 rows are missing, so the difference is trivial. We can just drop the resulting NAs.

ggplot(diet_preference, aes(pref.similarity, diet.similarity, color = level)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm") +
  facet_wrap(~site)

ggplot(diet_preference, aes(tongue.diff, pref.similarity)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm") +
  facet_wrap(~site)

ggplot(diet_preference, aes(tongue.diff, diet.similarity, color = level)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm") +
  facet_wrap(~site)
```

### 2021-04-20 code/ms2_niche_overlap.Rmd
switched to proportional visitation data for null models

---
title: "visitation_bias"
author: "Doug Sponsler"
date: "12/14/2020"
output: html_document
---
# Preparations
## Packages 
```{r, warning=FALSE, include=FALSE, echo = FALSE}
library(bipartite)
library(mgcv)
library(mgcViz)
library(modelr)
library(vegan)
library(ggvegan)
library(ggrepel)
library(tidymv)
library(patchwork)
library(tidymodels)
library(tidyverse)
library(lubridate)

rstan::rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

@@

# Visitation data
visitation <- read_csv("../data/processed_data/visitation.csv")
```

@@

## 3.3 Function for calculating niche overlap
```{r}
# This one averages visitation and then calculates niche overlaps
niche_overlapper <- function(x, criterion) {

@@

## 3.4 Calculate niche overlap pooled within site
```{r}
# Pooled across dates within level, site, and species pair
niche_overlap_gen <- niche_overlapper(niche_data, plant.genus) %>%
  mutate(level = rep("genus", n()),
         niche.overlap.genus = niche.overlap)

niche_overlap_fam <- niche_overlapper(niche_data, plant.family) %>%
  mutate(level = rep("family", n()),
         niche.overlap.family = niche.overlap)

niche_overlap_ktype <- niche_overlapper(niche_data, k.type.ss) %>%
  mutate(level = rep("morphotype", n()),
         niche.overlap.ktype = niche.overlap)

@@

# Collate across levels
niche_overlap <- bind_rows(niche_overlap_gen,
                           niche_overlap_fam, 
                           niche_overlap_ktype,) %>%
  mutate(level = factor(level, levels = c("family", "morphotype", "genus")),
         site = factor(site),
         management = factor(management, levels = c("none", "grasing", "mowing"))) %>%
  separate(sp.pair, c("sp1", "sp2"), sep = ":", remove = FALSE)

>>>>

---
title: "visitation_bias"
author: "Doug Sponsler"
date: "12/14/2020"
output: html_document
---
# Preparations
## Packages 
```{r, warning=FALSE, include=FALSE, echo = FALSE}
library(bipartite)
library(mgcv)
library(mgcViz)
library(modelr)
library(vegan)
library(ggvegan)
library(ggrepel)
library(tidymv)
library(patchwork)
library(tidymodels)
library(tidyverse)
library(lubridate)
```

@@

# Visitation data
visitation <- read_csv("../data/processed_data/visitation.csv")
visit_cover <- read_csv("../data/processed_data/visitation.csv")
```

@@

## 3.3 Function for calculating niche overlap
```{r}
# This one averages visitation and then calculates niche overlaps using 1 - Horn-Morisita distance
niche_overlapper <- function(x, criterion) {

@@

## 3.4 Calculate niche overlap pooled within site
```{r}
# Pooled across dates within level, site, and species pair
niche_overlap_gen <- niche_overlapper(niche_data, plant.genus) %>%
  mutate(level = rep("genus", n()))

niche_overlap_fam <- niche_overlapper(niche_data, plant.family) %>%
  mutate(level = rep("family", n()))

niche_overlap_ktype <- niche_overlapper(niche_data, k.type.ss) %>%
  mutate(level = rep("morphotype", n()))

@@

# Collate across levels
niche_overlap <- bind_rows(niche_overlap_gen,
                           niche_overlap_fam, 
                           niche_overlap_ktype) %>%
  mutate(level = factor(level, levels = c("family", "morphotype", "genus")),
         site = factor(site),
         management = factor(management, levels = c("none", "grasing", "mowing"))) %>%
  separate(sp.pair, c("sp1", "sp2"), sep = ":", remove = FALSE)
